---
Title: Lecture 6 Activities
Author: Julia Schedler
---

# Setup

## Part 0

-   Create a folder somewhere on your computer called "Lecture6".

-   Open R studio, and create a .qmd File and save inside the "Lecture6" folder as "Yourname_Lecture6.qmd".

-   Toggle into "Source" mode (upper left corner)

-   Copy the content on This Page and then toggle into visual mode

-   Click the render button. Does what you get look like what's on this page?

# Accounting for Trends

## Part 1

1.  Download the data `time_series.csv` from Canvas and create a sub-folder in your `Lecture6` folder called `Data`. Read in the data. Extract `y1`, the first time series, and name it `x_t`.

```{r}
library(astsa)
time_series <- read.csv("../Data/time_series.csv")
x_t <- time_series$y1
```

2.  Plot the data

```{r}
tsplot(x_t)
```

3.  In separate plot, again plot the time series and also plot a moving average smoother over the time series (use any $p$ you'd like).

```{r}
tsplot(x_t)

ma_x_t <- stats::filter(x_t, filter =rep(1/5,5))

lines(ma_x_t, col = "blueviolet")
```

3.  Detrend the time series with respect to the moving average estimate and plot the de-trended time series.

```{r}
detrended_ma <- x_t - ma_x_t
tsplot(detrended_ma)

```

4.  In a separate plot, again plot the time series and also plot a simple linear regression line using 1:length(x) where x is the time series you are analyzing.

```{r}
time_index <- 1:length(x_t)
lm_x_t <- lm(x_t~time_index)
tsplot(x_t)
reg_x_t <- lm_x_t$fitted.values
lines(reg_x_t, col = "turquoise", lwd = 2)
```

5.  Detrend the time series with respect to the regression and plot the detrended series.

```{r}
detrended_reg <- x_t - reg_x_t
tsplot(detrended_reg)
```

6.  Difference the time series and plot it in yet another plot.

```{r}
differenced <- c(diff(x_t), NA)
tsplot(differenced)
```

7.  `Run par(mfrow = c(2,3))` and then re-run the plotting code for all 6 plots you just made in steps 1-6.

```{r}
par(mfrow = c(2,3))
# number 1
tsplot(x_t)
  
## number 2
ma_x_t <- stats::filter(x_t, filter = rep(1/5, 5), method = "convolution")
tsplot(x_t)
lines(ma_x_t, col = "blueviolet", lwd = 2)
  
# number 3
detrended_ma <- x_t - ma_x_t
tsplot(detrended_ma)
  
# Number 4
time_index <- 1:length(x_t)
lm_x_t <- lm(x_t~time_index)
tsplot(x_t)
reg_x_t <- lm_x_t$fitted.values
lines(reg_x_t, col = "turquoise", lwd = 2)
  
# number 5
detrended_reg <- x_t - reg_x_t
tsplot(detrended_reg)

# Number 6
differenced <- c(diff(x_t), NA)
tsplot(differenced)
```

7.  Describe how the time series changed for each of the detrended series and also the differenced series. Do they appear stationary? (Make a table to organize the information.

| Processing           | Change     | Stationary?                                                                                                                      |
|---------------|-------------------------|-------------------------------------------------|
| x_t                  | \-         | Yes                                                                                                                              |
| Detrend (MA)         | Negligible | Yes                                                                                                                              |
| Detrend (Regression) | Negligible | Yes, also, the original regression line looks like it has a slope of about 0 and there don't appear to be any seasonal patterns. |
| Difference           | Negligible | Yes                                                                                                                              |

## Part 2

Close Rstudio and re-open. Repeat Part1 on `y2` . To make your life easier, toggle into visual mode and copy everything from part 1 in source mode, scroll down under part 2, and then toggle back to Visual mode.

1.  Read in the data. Extract `y2`

```{r}
#library(astsa)
#time_series <- read.csv("../Data/time_series.csv")
x_t <- time_series$y2
```

2.  Plot the data

```{r}
tsplot(x_t)
```

3.  In separate plot, again plot the time series and also plot a moving average smoother over the time series (use any $p$ you'd like).

```{r}
tsplot(x_t)

ma_x_t <- stats::filter(x_t, filter =rep(1/5,5))

lines(ma_x_t, col = "blueviolet")
```

3.  Detrend the time series with respect to the moving average estimate and plot the de-trended time series.

```{r}
detrended_ma <- x_t - ma_x_t
tsplot(detrended_ma)

```

4.  In a separate plot, again plot the time series and also plot a simple linear regression line using 1:length(x) where x is the time series you are analyzing.

```{r}
time_index <- 1:length(x_t)
lm_x_t <- lm(x_t~time_index)
tsplot(x_t)
reg_x_t <- lm_x_t$fitted.values
lines(reg_x_t, col = "turquoise", lwd = 2)
```

5.  Detrend the time series with respect to the regression and plot the detrended series.

```{r}
detrended_reg <- x_t - reg_x_t
tsplot(detrended_reg)
```

6.  Difference the time series and plot it in yet another plot.

```{r}
differenced <- c(diff(x_t), NA)
tsplot(differenced)
```

7.  `Run par(mfrow = c(2,3))` and then re-run the plotting code for all 6 plots you just made in steps 1-6.

```{r}
par(mfrow = c(2,3))
# number 1
tsplot(x_t)
  
## number 2
ma_x_t <- stats::filter(x_t, filter = rep(1/5, 5), method = "convolution")
tsplot(x_t)
lines(ma_x_t, col = "blueviolet", lwd = 2)
  
# number 3
detrended_ma <- x_t - ma_x_t
tsplot(detrended_ma)
  
# Number 4
time_index <- 1:length(x_t)
lm_x_t <- lm(x_t~time_index)
tsplot(x_t)
reg_x_t <- lm_x_t$fitted.values
lines(reg_x_t, col = "turquoise", lwd = 2)
  
# number 5
detrended_reg <- x_t - reg_x_t
tsplot(detrended_reg)

# Number 6
differenced <- c(diff(x_t), NA)
tsplot(differenced)
```

7.  Describe how the time series changed for each of the detrended series and also the differenced series. Do they appear stationary? (Make a table to organize the information.

| Processing           | Change     | Stationary?                                            |
|----------------|------------------|--------------------------------------|
| `y2`                 | \-         | Yes                                                    |
| Detrend (MA)         | Negligible | Yes                                                    |
| Detrend (Regression) | Negligible | Yes (regression line looks to have a slope of about 0) |
| Difference           | Negligible | Yes                                                    |

## Part 3

Create a new folder in your `Lecture6` Directory called `Code`. Download the file `models.R` from Canvas and save it in the `Code` folder.

1.  Inspect the code in `models.r`. What is it doing?
2.  Run the function `models` on `y3.`

```{r}
source("../Code/models.R")

models(time_series$y3, ma_order = 5)

```

3.  Describe how the time series changed for each of the detrended series and also the differenced series. Do they appear stationary? (Make a table to organize the information.)

| Processing           | Change     | Stationary?                                            |
|------------------|------------------|-------------------------------------|
| `y3`                 | \-         | Yes                                                    |
| Detrend (MA)         | Negligible | Yes                                                    |
| Detrend (Regression) | Negligible | Yes (regression line looks to have a slope of about 0) |
| Difference           | Negligible | Yes                                                    |
