---
title: "Lecture 5 - Draft"
author: Julia Schedler
format: 
  html: 
    code-fold: show
slide-number: true
execute: 
  eval: false
scrollable: true
---

## [Activity 1: Detrending a commodity (Example 3.7)]{.r-fit-text} {.smaller}

-   Given the code to generate the plot with the trend line, how would you view the equation of the trend line?

```{r}
#| eval: true

```
-   Visualize the de-trended series. Does it appear stationary?

```{r}
#| eval: false

## your code here
```
-   Compute the acf of the salmon series and the detrended series. What do you notice?
```{r}
#| eval: false 

## your code here


```


## [Activity 2: "assuming we are able to estimate it"]{.r-fit-text}

-   Look at pages 37-41 of the textbook
-   what is "it" in this context? (what are we estimating?)
-   If this is review, where did you first see these ideas?
-   Put a dot on the math anxiety rating distribution on the back board


## [Activity 3: Simulating a random walk and then differencing it]{.r-fit-text}
1. Simulate a random walk with no drift and plot it. 

```{r}
#| eval: false

## Simulate random walk w/ drift
## your code here 
```
2. Use the `diff` function to difference the simulated series. Plot the result.
```{r}
#| eval: false
# your code here
```
3. Does this series appear stationary? How do you know? 
4. Visualize the ACF of the differenced series. Does it look like white noise?
```{r}
#| eval: false
# your code here
```


## [Activity 4: Differencing Salmon Prices]{.r-fit-text}
1. Compute and plot the differenced salmon series. 

```{r}
#| eval: false
# your code here
```

2. Does the series appear stationary?
3. Visualize the acf of the differenced series. Does it look like white noise?
```{r}
#| eval: true
```

## [Activity 5: Comparing Differencing and Detrending]{.r-fit-text}
Compare the Acfs of the differenced and detrended salmon series. What do you notice?
```{r}
#| eval: false

# your code here
```
