---
title: Lecture 9
author: Julia Schedler
format: html
---

## Announcements{.smaller}

- Office hours:
  - today: 4-6pm
  - tomorrow: 9:30-11 and 2:30-4 (note different than announcement)
  
- Assignment 3: 
  - Solutions posted tonight
  - ask any questions you'd like in class
  - Hopefully have grades tonight, for sure tomorrow

## Time Series Data Analysis Process

- What should it be?
- Scientific method?
- What if I don't need a scientific level of rigor?

## [Scientific method (according to wikipedia)]{.r-fit-text}{.smaller}

An iterative, pragmatic scheme of the four points above is sometimes offered as a guideline for proceeding:

1. Define a question
2. Gather information and resources (observe)
3. Form an explanatory hypothesis
4. Test the hypothesis by performing an experiment and collecting data in a reproducible manner
5. Analyze the data
6. Interpret the data and draw conclusions that serve as a starting point for a new hypothesis
7. Publish results
8. Retest (frequently done by other scientists)

The iterative cycle inherent in this step-by-step method goes from point 3 to 6 and back to 3 again.


## [Scientific method (according to ChatGPT)]{.r-fit-text}{.smaller}

1. Observation
2. Question
3. Hypothesis
4. Experiment
5. Data Collection and Analysis
6. Conclusion
7. Report Results
8. Replication

**note: interesting the first two steps are flipped!**

## What about a non-scientific context?

- Do businesses do science?
- Methods of iterative inquiry are not owned by scientists

## [Business Problem-solving process(chatGPT)]{.r-fit-text}{.smaller}

1. Identify the Problem (similar to observation)
2. Define Objectives/Goals (related to formulating a question)
3. Generate Hypotheses/Solutions (like hypothesis formation)
4. Test/Implement Solutions (similar to experimentation)
5. Analyze Results (equivalent to data collection and analysis)
6. Make Decisions (conclusion phase)
7. Monitor and Review (parallels replication)

## A personal note

What does **data driven** mean? 

- I thought scientific
- I thought wrong :(
- I got very frustrated and considered getting an MBA just so I could understand the motivations of management
- I gate-kept the term research on a company-wide zoom meeting
- I got laid off and became a research scientist :)

**Defining goals clearly for analytic thinkers is very important. Clarification questions should not be intended or received as hostile. Alas.**

## Activity 1{.smaller}

Here are some time series data analysis tools we have learned in class. Where do they fit in the scientific method?

- time series plot
- autocorrelation plot
- lag plot
- trend estimation
- detrending 
- differencing
- writing out a mathematical model
- optimizing plots (adding good axis labels, etc)
- ask a time series research question

## Activity 1 Solutions

(using chatGPT list)

- time series plot: **1, 5, 6, 7**
- autocorrelation plot: **1, 5**
- lag plot: **5**
- trend estimation: **5, 6, 7**
- detrending: **5**
- differencing: **5**
- writing out a mathematical model: **5, 7**
- optimizing plots (adding good axis labels, etc): **7**
- ask a time series research question: **2, 3**


## Activity 2
What are some things we have *not* done?


## Activity 2 Solutions
What are some things we have *not* done?
- replicate results
- test a specific hypothesis (well, we kind of have)
- report/publish results
- experiment / collect data

## Activity 3: Applying the data analysis process

1. Create a .qmd file
2. Add headers for each step of the scientific method or business problem-solving process
3. Apply the appropriate tools at each step of the process for the AirPassengers data.

## 1. Observation

```{r}
#| echo: true
#| collapse: true
#| code-fold: true
library(astsa)
?AirPassengers
summary(AirPassengers)
tsplot(AirPassengers)
```

## 2. Question {.smaller}

- Are the number of air passengers increasing over time? If so, how fast?
  - Is there an increasing trend? What is the form of the trend?
- Is there a seasonality to the monthly number of air passengers? When are the peaks?
  - Is there a seasonal trend? What is the period? When are the peaks?
- What is a plausible range of values for the predicted number of passngers for the next month?
  - Can we make a forecast? If so, how good do we expect this forecast to be?

## 3. Hypothesis

- There is an increasing linear trend.
- There is a seasonal component.
- We can accurately model the time series using tools that allow forecasting.

## 4. Experiment

- Not really applicable here...

## 5. Data collection and analysis

- Data already collected! In practice, we may need to clean it.
- Exploratory data analysis: decompose?

```{r}
plot(decompose(AirPassengers)) ## hmm
``` 

## 5. Data collection and analysis

- Data already collected! In practice, we may need to clean it.
- Exploratory data analysis: decompose?

```{r}
AP_decomp <- decompose(AirPassengers, type = "multiplicative")
plot(AP_decomp) ## hmm
``` 

## 5. Data Collection and analysis
- Doesn't look quite right... maybe we should talk to the senior analyst about a more advanced modeling technique?

```{r}
acf(AP_decomp$random, na.action = na.pass)
```
## 5. Data Collection and analysis
```{r}
lag1.plot(AirPassengers, 12) ## hmmm
```
## 5. Data Collection and analysis
```{r}
lag1.plot(AP_decomp$random[!is.na(AP_decomp$random)], 12, ) ## hmmm
```
Linear fit looks ok, but still some structure in the residuals.

## 6. Conclusion{.smaller}

- It appears there is some trend and seasonality in the number of monthly air passengers, 
- but we need some more advanced modeling tools to fully capture it.
- Current model:

$$
x_t = \mu_t\cdot s_t  + y_t
$$
  
  - $x_t$ is the data
  - $y_t$ is (ideally) white noise
  - $\mu_t$ is an increasing trend (estimated by moving average)
  - $s_t$ is a seasonal component
  - we make a *multiplicative* assumption
  
## A peep ahead{.scrollable}

```{r}
#| echo: true

plot(AirPassengers, type = "l")
points(AP_decomp$seasonal*AP_decomp$trend, col = "blueviolet", type = "l")

AP.hw <- HoltWinters(AirPassengers, seasonal = "mult")
plot(AP.hw)
AP.predict <- predict(AP.hw, n.ahead = 4*12)
plot(AP.hw, type = "l", col = "red", xlim = c(1950, 1965), ylim = c(100, 800))
points(AirPassengers, type = "l")
points(AP.predict, type = "l", lty = 2, col = "red") 
acf(AirPassengers - fitted(AP.hw)[,1])
```
  
## Report results

- Paper write up
- Slide presentation
- Video summary

Lots of ways to report results...

## Questions on anything!

