---
title: Assignment 4 Solutions
author: Julia Schedler 
---

Remember, you are allowed to use AI on this assignment. However, you must provide a link to a transcript. If you are working with a classmate, please include the name of that classmate.

## Snowshoe hare furs

Consider the number of snowshoe hare furs sold by the Hudson Bay Company between 1845 and 1935. You can access the dataset, `pelt`, in `R` by loading the `fpp3` package.

1. Produce a well-labeled time series plot of the Snowshoe Hare furs. 

```{r}
#| label: p1-1
#| collapse: true
#| message: false
library(fpp3)
library(astsa)

tsplot(x = pelt$Year, y = pelt$Hare, main = "Snowshoe Hare Pelts", 
       ylab = "Number of Pelts", 
       xlab = "Year", 
       type = "b", 
       pch = 17)
```

2. Describe the plot produced in number 1.

The snowshoe hare pelt does not appear to have an increasing/decreasing trend nor a seasonal trend-- the peaks and valleys are not equally spaced. There is clearly some time dependence, however, because the series does not look like white noise.

3. Make an acf and pacf of the data. Suggest a model for the snowshoe hare fur *based on the acf/pacf*.

```{r}
par(mfrow = c(1,2))
acf1(pelt$Hare)
pacf(pelt$Hare)
```
The acf appears to taper off and the pacf appears to spike and then cut off at a lag of 2. This suggests an AR(2) process.

4. Fit the model you suggested in number 3.

```{r}
ar2_hare <- sarima(pelt$Hare, p = 2, d = 0, q = 0, details = FALSE)

```

5. Write the equation of the model you fit in number 4. 
If $x_t$ is the number of snowshoe hare pelts sold at time $t$, then 

$$
x_t = 44638.3226 
$$

6. Provide and interpret 3-5 diagnostic plots or tests and comment on whether the model fits well.

4. **By hand**, calculate forcasts for the years 1936-1939.

5. Using R, calculate forecasts for the years 1936-1939. 

6. Plot the forecasts on the same plot as the data, and interpret the forecast.

7. Use the automatic algorithm in the `ARIMA()` function to fit a model to the snowshoe hare time series.

8. Compare the models and provide a recommendation on choice of model (use a 2 pieces of "evidence" for your recommendation). Make sure to mention differences in forecasts, if any.
