---
title: "Lecture 4"
author: Julia Schedler
format: 
  html: 
    code-fold: show
slide-number: true
scrollable: true
filters: 
  - timer
---

::: {.callout-note collapse="true" appearance="simple" icon="false"}
## Slides with Activity Solutions

[View slides in full screen](Slides/Lecture4/){target="_blank"}
:::

::: {.callout-note collapse="true" appearance="simple" icon="false"}
## Link to Notes Template
[Lecture4Template.qmd](ActivityTemplates/Lecture4Template.qmd)
:::

## Recap

::: hidden
$$
\newcommand\E{{\mathbb{E}}}
$$
:::

-   Decomposing time series
-   Stationarity (theoretically and with data)
-   Some activities

## Today

-   Finish up activities from Lecture 3
-   Trend stationarity
-   Visualizing autocovariance (third attempt)
-   Detrending
-   "Office hours"

## Reminders

-   Syllabus participation policy
-   Assignments: Going forward, must submit **rendered pdf** of code portion. (if you want to be nice to me, do this for Assignment 2, but starts with Assignment 3)
-   Late quizzes: Going forward: Email me ahead of time, otherwise it's a 0
-   Exam details: No use of computer, code will be covered but basic, notes sheet is allowed, practice test will be provided

## [Activity 1: Export Price of Salmon (Example 3.1)]{.r-fit-text} {.smaller}

```{r}
#| label: activity-1
#| echo: true
#| collapse: true
#| code-fold: true
library(astsa)

summary(fit <- lm(salmon~time(salmon), na.action=NULL))
tsplot(salmon, col=4, ylab="USD per KG", main="Salmon Export Price")
abline(fit)

```

## [Activity 1: Export Price of Salmon (Example 3.1)]{.r-fit-text} {.smaller}

1.  Does this time series appear stationary?

2.  The (mathematical) equation in the book for the trend line above is:

    $$
    x_t = \beta_0 + \beta_1z_t + w_t, z_t = 2003\frac{8}{12}, 2001\frac{8}{12}, \dots, 2017\frac{5}{12}
    $$

    -   There is a typo in this equation. Correct the typo. (hint: examine to the first few entries of `time(salmon)`

    -   Why are there fractions of the year? Explain what the fractional values mean and describe how they appear in the data set within R.

3.  Interpret the estimate of the slope.

## [Activity 2: Trend Stationarity (Example 2.19)]{.r-fit-text} {.smaller}

Consider the time series model $$x_t = \beta t + y_t$$ Assume $y_t$ is stationary with mean function $\mu_y$ and and autocovariance function $\gamma_y(h)$

-   Compare this equation to the regression equation in the last example.
-   What are the mean function and autocovariance function of $x_t$?

## Trend stationarity model

A time series which is nonstationary in the mean but is stationary in the autocovariance is sometimes called **Trend stationarity**.

-   I'm actually not sure if it just refers to linear trends?? I'll ask people at my conference

# Visualizing (possibly nonstationary) Autocorrelation functions

## Visualizing the autocovariance for trend stationarity

```{r}
#| label: show-error-dists
#| code-fold: true

library(ggplot2)
set.seed(807)
t <- seq(1, 10, 1)
x_t <- 0.5*t 
#x <- x - mean(x)
#y <- y - mean(y)

df <- data.frame(t, x_t)

# For every row in `df`, compute a rotated normal density centered at `y` and shifted by `x`
curves <- lapply(seq_len(NROW(df)), function(i) {
  mu <- df$x_t[i]
  range <- mu + c(-1.5, 1.5)
  seq <- seq(range[1], range[2], length.out = 100)
  data.frame(
    t = -1 * dnorm(seq, mean = mu, sd = 0.5) + df$t[i],
    x_t = seq,
    grp = i
  )
})
# Combine above densities in one data.frame
curves <- do.call(rbind, curves)

new.x = seq(from = 1, to = 10, by = .1)
new.y = .5*new.x
trend_line <- data.frame(x = new.x,
                       y = new.y)
ggplot(df, aes(t, x_t)) +
  geom_point(col = "blueviolet", pch = 17) +
  #geom_line() +
  # The path draws the curve
  geom_path(data = curves, aes(group = grp)) +
  geom_line(data = trend_line, aes(x=x,y=y), col = "blueviolet") +
  lims(y = c(-2,10)) +
  scale_x_continuous(breaks = seq(1, 10, by = 1)) +
  theme_minimal() + 
  theme( # remove the vertical grid lines
           panel.grid = element_blank() ,
           # explicitly set the horizontal lines (or they will disappear too)
           panel.grid.major.x = element_line( size=.1, color="black" )) +   
  geom_rect(aes(xmin = 1.1, xmax = 2.1, ymin = -1, ymax = 3), fill = NA, col = "blue")+   
  geom_rect(aes(xmin = 7.1, xmax = 8.1, ymin = 2, ymax = 6), fill = NA, col = "magenta")
  # The polygon does the shading. We can use `oob_squish()` to set a range.
  #geom_polygon(data = curves, aes(y = scales::oob_squish(y, c(0, Inf)),group = grp))
```

## [Simulate many time series from the trend stationarity model]{.r-fit-text} {.smaller}

```{r}
#| code-fold: true

set.seed(807)
Time <- 10
n_sim <- 100
all_series = matrix(rep(NA, times = n_sim*Time), nrow = Time)
for(i in 1:n_sim){
  cs = 0.25*1:Time        # same thing 
  w  = rnorm(Time + 50,0,1)
  #w_dep = stats::filter(w, filter = rep(1/3,3))[2:(Time + 1)]
  #all_series[,i] <- cs + w_dep
  all_series[,i] <- cs + w[2:(Time+1)]
  #names(all_series)[i] <- paste("sim_", i)
  
}
fit <- lm(all_series[,1]~time(all_series[,1]), na.action=NULL)
tsplot(all_series[,1], main = "One simulated Time series", type = "b", col = c("black", "blue", rep("black", times = 5), "magenta", "black", "black"), pch = 16, cex = 3)
abline(fit)

```

## [Simulate many time series from the trend stationarity model]{.r-fit-text} {.smaller}

```{r}
#| code-fold: true
par(mfrow = c(2,2))
for(i in 1:4){
  fit <- lm(all_series[,i]~time(all_series[,i]), na.action=NULL)
  tsplot(all_series[,i], type = "b", col = c("black", "blue", rep("black", times = 5), "magenta", "black", "black"), pch = 16, cex = 3, ylim = c(-3,6))
  abline(fit)

}
```

## [Simulate many time series from the trend stationarity model]{.r-fit-text} {.smaller}

```{r}
#| label: trend-stationarity-simulate
#| echo: true
#| collapse: true
#| code-fold: true

tsplot(all_series, spaghetti = TRUE, main = "100 Simuated Trend Stationary Time Series", type = "b")
rect(xleft = 1.5, xright = 2.5, ybottom = -2, ytop = 4, border = "blue", lwd = 2)
rect(xleft = 7.5, xright = 8.5, ybottom = -.5, ytop = 5, border = "magenta", lwd = 2)

```

## [Marginal and Joint Distributions t=8 and s=2]{.r-fit-text} {.smaller}

```{r}
#| code-fold: true
i = 2
x_comp <- all_series[i,]
x_8 <- all_series[8,]

par(mfrow = c(2,2))
hist(x_comp, col = "blue", main = paste("Histogram of simulations at t=", i), xlab = paste("x_",i))
hist(x_8, col = "magenta", main = paste("Histogram of simulations at t=8"), xlab = "x_8")
plot(x_comp, x_8, col = "purple", pch = 16, main = "Joint Distribution")
abline(lm(x_8~x_comp))

```

## Marginal and Joint Distributions

```{r}
#| code-fold: true
library(ggplot2)
library(ggExtra)
x <- data.frame(x_comp, x_8)
# Save the scatter plot in a variable
p <- ggplot(x, aes(x = x_comp, y = x_8)) +
  geom_point(col = "purple") + xlim(-3,6) + ylim(-3,6)+ 
  geom_text(aes(x = 2, y = -1, label = paste("rho(",i,",8) = \n cor(x_",i,", x_8) = ", round(cor(x_comp, x_8),3)), size = 6)) + coord_fixed() 

# Arguments for each marginal histogram
ggMarginal(p, type = "density", adjust = 2,
           xparams = list(col = "blue", fill = "blue"),
           yparams = list(col = "magenta", fill = "magenta"))
```

## Do this for all possible combos of $s$ and $t$

[View slides in full screen](Slides/SimulateAutocov/){target="_blank"}

## Visualizing the correlations

```{r}
#| code-fold: true
cors <- apply(all_series, 1, function(x){cor(x, all_series[8,])})
hist(cors, breaks = seq(-1, 1, by = .1), main = "Correlation between simulations for x_8 and x_i")
```

## Are these all the correlations?

No, just pairwise with $x_8$. We could do all possible pairs:

```{r}
#| code-fold: true
library(GGally)
ggpairs(data.frame(t(all_series)))
```

## Distribution of all the correlations:

```{r}
#| code-fold: true
coords <- t(combn(10,2))
cors <- NULL
for(i in 1:nrow(coords)){
  cors <- c(cors, cor(all_series[coords[i,1],], all_series[coords[i,2],]))
}
cors <- c(cors, rep(1, times = 10))
hist(cors, breaks = seq(-1, 1, by = .1), main = "Correlation between simulations for x_s and x_t")
```

# 3D plots

## 3D version of histogram (includes $s-t$ plane)

```{r}
#| code-fold: true
coords <- expand.grid(1:Time, 1:Time)
names(coords) <- c("s", "t") 
coords$cor <- NA
coords$pval <- NA
cor_mat_theoretical <- cor_mat <- matrix(rep(NA, times = Time*Time), nrow = Time)
for(i in 1:nrow(coords)){
  out <- cor.test((all_series[coords[i,1],]), (all_series[coords[i,2],]))
  if(coords[i,1]==coords[i,2]){
    coords$cor[i] <- 1
    coords$pval[i] <- 0
    cor_mat[coords[i,1], coords[i,2]] <- 1
    cor_mat_theoretical[coords[i,1], coords[i,2]] <- 1
  }else{
    coords$cor[i] <- out$estimate
    coords$pval[i] <- out$p.value
    cor_mat[coords[i,1], coords[i,2]] <- out$estimate
    cor_mat_theoretical[coords[i,1], coords[i,2]] <- 0
    
  }
}

library(plotly)
plot_ly(coords,
        x= ~s, y=~t, z=~cor, 
        type = 'scatter3d', mode = "markers", size = .1)
```

## [Simulated auto correlation function $\hat{\gamma}_x(h)$ (with a blanket)]{.r-fit-text}

```{r}
#| code-fold: true
## "drape a blanket over it to see the pattern better"
fig <- plot_ly(z = ~cor_mat)
fig %>% add_surface()
```

## [Theoretical $\hat{\gamma}_x(h)$ (using the derived formula (with a blanket)]{.r-fit-text}

```{r}
#| code-fold: true
# Theoretical function
fig <- plot_ly(z = ~cor_mat_theoretical)
fig %>% add_surface()

```

# Detrending

## Detrending

If a process is trend stationary (nonstationary in the mean, but stationary in the variance), can we just subtract off the trend and get back a stationary time series?

Yes, and that's called **detrending**

## [Activity 3: Detrending a commodity (Example 3.7)]{.r-fit-text} {.smaller}

-   Given the code to generate the plot with the trend line, how would you view the equation of the trend line?
-   Visualize the de-trended series
-   Compute the acf of the salmon series and the detrended series. What do you notice?

## Next time

-   Cross-correlation and regression with multiple time series ($x_t$ on x-axis instead of $t$ on x-axis like with the salmon)
-   Activities and examples
-   Smoothing
